---
title: "Reproducible Research"
author: "John Atanga"
date: "7/7/2021"
output: 
  html_document: 
    keep_md: yes
---
## This is my document for the Assignment
## Load the data, in this case one call the read.csv():
## Process/ Transform the data inti a format suitable  for the analysis 
```{r dataload, echo=TRUE}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) { 
  temp <- tempfile()
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
  unzip(temp)
}
activity <- read.csv("activity.csv")
head(activity)
```
## I will like to view some information about variables
```{r }
str(activity)
```
## The following variables are included in this dataset:
## 1. **steps** : Number od steps taken in 5-minutes interval( missing values are coded as NA)
## 2. **date**  : The date on which the measurement was taken in YYYY-MM-DD format
## 3. **interval** : Identifier for the 5-minute interval in which measurement was taken.

## I will Process and transform the data  into a format that is suitable for the analysis

## **What is the mean number of steps taken per day?**
## For this part of the assignment one can ignore  the  missing values in the dataset. 1. I will calculate the total number of steps taken per day.
```{r }
totalStepsByDay <- aggregate(steps~date, activity, sum)
```
## 2. I will make a histogram of the total number of steps taken each day

```{r }
hist(totalStepsByDay$steps, xlab = "Class of Total Number of Steps per day", ylab = "Number of Days", main = "Total Number of Steps taken each day")

```

## One can also calaculate and report the mean and median of the total number of steps taken per day

```{r  mmedia, echo=TRUE}
mean_raw <- mean(totalStepsByDay$steps)
mean_raw
```
## Mean number of number of steps taken per day is 1.076618910^{4}
```{r }
median_raw <- median(totalStepsByDay$steps)
median_raw
```
## Median number of steps taken per day is 10765

## **What is the average daily activity pattern?**
## 1. One can make a time series plot ( type = "l") of the 5-minutes interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
## The average number of steps taken:

```{r steps, echo=TRUE}
averageStepsbyInterval <- aggregate(steps~interval, activity, mean)

```

## Time series plot of the average number of steps taken:

```{r }
with(averageStepsbyInterval, plot(interval, steps, type = "l"))
```

## 2. One can determine which 5-minutes interval, on average across all the days in the dataset, that contains the maximum number of steps, is a good question, ready to go

```{r }
averageStepsbyInterval[which.max(averageStepsbyInterval[,2]), 1]
```
## **Imputing missing values**
## There were a number of days/intervals with missing values - coded as NA. thus presenting bias.
## 1. One can calculate and report the total number of dmissing values in the dataset, ie NAs

```{r }
missingIndex <- is.na(activity[, 1])
```
## There are 2304 missing values in the dataset
## 2. Device a strategy for filling in all of the missing values in the dataset. use the mean / median for that day or for 5-minutes interval. 
## I used the mean number of steps perinterval to fill missing values
## Finding the mean number of steps per interval
```{r }
m <- mean(averageStepsbyInterval$steps)
```

## 3. One can create a new dataset that is equal to the original dataset but wiht the missing data filled in 
Imputing missing values  with M=37.3825996

```{r }
activity1 <- activity
activity1[missingIndex, 1] <- m

```

## Make a histogram of the total number of steps taken each day and calculate and report the mean an d median number  of steps taken per day. are they different, what is the impact
## Finding the total number of steps each day after missing values are imputed and making histogram

```{r dailysteps, echo=TRUE}
   totalStepsByDay1 <- aggregate(steps~date, activity1, sum)
   hist(totalStepsByDay1$steps, xlab = "Class of Total Number of Steps per day", ylab = "Number of Days", main = "Number of Steps taken each day after missing values are imputed")

```

## To calculate the mean and median total number of steps per day, first find the total number of steps per day.

```{r }
totalStepsByDay1 <- aggregate(steps~date, activity1, sum)
```
```{r }
mean_afterImput <- mean(totalStepsByDay1$steps)
mean_afterImput
```
## Mean number of steps taken per day is 1.076618910^{4}
```{r }
median_afterImput <- median(totalStepsByDay1$steps)
median_afterImput
```
## Median number of steps taken per day is 1.076618910^{4}
## By imputting the missing values by the mean number of steps per interval, there is no differnce in mean before and after imputing that is surprising. The median has changed a little bit.

## ** Are there differnces in activity  patterns between weekdays and weekends?**
## Using the weekdays() and the dataset  with the filled  missing values.
## 1. Create a new factor variable in the dataset with two levels - 'Weekdays" and "Weekends"  indicating whether a given date is a weekend or weekday.

```{r wkd, echo=TRUE}
 activity1$date <- as.Date(activity1$date)
 library(dplyr)
```
```{r weeky, echo=TRUE}
 activity2 <- activity1%>%
  mutate(dayType= ifelse(weekdays(activity1$date) == "Saturday" | weekdays(activity1$date) == "Sunday", "Weekend", "Weekday"))
head(activity2)
```
## 2. Make panel plot  containing a time series  plot ie type = "l" of the 5-minutes interval and the average number of steps taken, averaged across all weekdays days and weekends days

```{r tmseries, echo=TRUE}
 averageStepByDayTypeAndInterval <- activity2 %>% group_by(dayType, interval) %>%
 summarize(averageStepByDay=sum(steps))
 
 head(averageStepByDayTypeAndInterval)

```
```{r lat, echo=TRUE}
 library(lattice)
 with(averageStepByDayTypeAndInterval, xyplot(averageStepByDay ~ interval | dayType, type = "l", 
 main = "Total Number of Steps within Intervals by dayType", xlab = "Daily Intervals", ylab = "Average Number of Steps"))
```



